---
title: "Case control"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("install_packages.r")
require(knitr)
require(ggplot2)
require(dplyr)
require(GGally)
```

## Case-control sampling and logistic regression ####


Often, in binary dependent variables, "success" cases are rare and we take them all into our sample.

Rare diseases vs. population, e-shop purchase vs. visits 

All "successful" cases plus approx. 5-times that number of controls is usually sufficient for Logit
 provided random selection


## Example: Myocardial infarction (MI) in South Africa

The variables are

|   variable         | description                      |
|-----------------|-----------------------------------------------|
|   sbp           | systolic blood pressure;                      |
|     tobacco     |      cumulative tobacco (kg);                 |
|     ldl         |  low densiity lipoprotein   cholesterol;      |
|     adiposity   |  a measure of fat;                            |
|     famhist     |   family history of heart disease   (Present  |
|     obesity     |    BMI                                        |
|     alcohol     |    current alcohol consumption;               |
|     age         |      age at onset;                            |
|     chd         |        response  coronary heart disease (MI). |

```{r}
rm(list=ls())
heart <- read.csv("dta/SouthAfrica.csv")
```

Sample of adult males in a heart-disease high-risk region of the Western Cape, South Africa, approx. 1980.

There are roughly two controls per case of CHD.

The prevalence rate of MI is 5.1 % in the region.


```{r}
head(heart, 10)
summary(heart)
summary(as.factor(heart$chd)) 
plot(heart[ , 1:8], cex=0.8, pch=16, col=as.factor(heart$chd))
```

```{r, warning=FALSE, message=FALSE}
ggpairs(data=cbind(heart[,c("age","tobacco","alcohol","obesity","famhist")],chd=as.factor(heart$chd)), 
        aes(colour = chd), 
        upper = list(continuous = "density", combo="box"))
```

```{r}
ggplot(data=heart)+
  geom_boxplot(aes(as.factor(chd),obesity))+
  facet_wrap(~cut(heart$age,7))
```


Our goal is to identify relative strengths and directions of risk factors.

## Logistic regression

```{r}
heartfit <- glm(chd~., data = heart , family = binomial)
summary(heartfit)
```

With case-control samples, we can estimate the regression
parameters beta(j) accurately - if our model is correct.

The constant term beta(0) is biased.
.. but may be corrected relatively easily

```{r}
heart.coeffs <- heartfit$coefficients
heart.coeffs
```

```{r}
pi.pop <- 0.051 # percentage in population
pi.smpl <- 160/462 # percentage in sample
```

case control adjustment formula for the (Intercept)

$$\hat{\beta}_0^\ast = \hat{\beta}_0 + \log\frac{\pi}{1-\pi}-\log\frac{\tilde{\pi}}{1-\tilde{\pi}}$$

```{r}
(beta.0.pop <- heart.coeffs[1] + log(pi.pop/(1-pi.pop)) - log(pi.smpl/(1-pi.smpl)))
```


```{r}
heart.coeffs[1] <- beta.0.pop
heart.coeffs
length(heart.coeffs) # Intercept + 8 variables
```


Lets calculate MI probability for an "average" SA male with family history of MI
and for "the same" male without family history of MI


```{r}
attach(heart)
heart2 <- rbind(heart, list(mean(sbp), mean(tobacco), mean(ldl), mean(adiposity), "Present", mean(obesity),
                 mean(alcohol), mean(age), NA))
heart2 <- rbind(heart2, list(mean(sbp), mean(tobacco), mean(ldl), mean(adiposity), "Absent", mean(obesity),
                            mean(alcohol), mean(age), NA))
tail(heart2)
```



Biased predictions, based on the sample "success" prevalence (beta 0 is biased)

```{r}
predict(heartfit, newdata=heart2[463:464, ], type="response")
```

## Case-control corrected predictions

```{r}
heart3 <- heart2 # We prepare yet another dataframe - for illustration and comparison
heart3$famhist <- 0  # with dummy variables, because we cannot multiply
heart3$famhist[heart2$famhist == "Present"] <- 1 # dummy variable instead of factor
(Log.Odds1 <- heart.coeffs[1] + sum(heart3[463, 1:8]*heart.coeffs[2:9]))
(Log.Odds2 <- heart.coeffs[1] + sum(heart3[464, 1:8]*heart.coeffs[2:9]))
```

Expected probability of MI:

```{r}
Average.with.famhistory.Yes <- exp(Log.Odds1)/(1+exp(Log.Odds1))
Average.with.famhistory.No <- exp(Log.Odds2)/(1+exp(Log.Odds2))
names(Average.with.famhistory.Yes) <- "Expected MI probability: average male WITH family history of MI"
names(Average.with.famhistory.No) <- "Expected MI probability: average male WITHOUT family history of MI"
```

```{r}
Average.with.famhistory.Yes
Average.with.famhistory.No
```

