---
title: "Choropleths"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
require(knitr)
require(dplyr)
require(plotly)
require(ggplot2)
require(gridExtra)
library(rgeos) 
library(rgdal)
require("SmarterPoland")
require("reshape") 
require(RColorBrewer)
```

## Advanced R mapping: Choropleths / infomaps 
   ... using ggplot and macroeconomic data


  
   Choropleth for 
   GDP per capita in % of EU average
   Austria, Czech Republic, Germany, Hungary, Poland, Slovakia


```{r}
# Skip this section, if geo-data are already downloaded
download.file("http://ec.europa.eu/eurostat/cache/GISCO/geodatafiles/NUTS_2010_60M_SH.zip", 
              destfile = "NUTS_2010_60M_SH.zip")
# unzip to SpatialPolygonsDataFrame
unzip("NUTS_2010_60M_SH.zip")
```
 
## Step 1 - Preparation of "mappning data"

```{r}
map <- readOGR(dsn = "./NUTS_2010_60M_SH/data", layer = "NUTS_RG_60M_2010")
# Retrieve Austria, Czech Republic, Germany, Hungary, Poland, Slovakia
# polygon data at all NUTS levels.
# Now, we create a SpatialPolygonsDataFrame object for the selected countries
# STAT_LEVL_ is NUTS2 regions only, we only need level 2
map$NUTS_ID[1:4] #first two letters are country code, substr(x, start, stop)
CE_map <- map[substr(map$NUTS_ID,1,2) %in% c("AT","CZ","DE","HU","PL","SK") & map$STAT_LEVL_ == 2,]
CE_map.borders <- map[substr(map$NUTS_ID,1,2) %in% c("AT","CZ","DE","HU","PL","SK") & map$STAT_LEVL_ == 0,]
```

## Step 2 - Download & preparation of the macroeconomic data

 ... GDP Per Capita in EUR, percentage of EU average
 ... data source: Eurostat (consistent NUTS-based observations)
 
 
 
```{r}
require("SmarterPoland") # install.packages("SmarterPoland")
require("reshape")
# Download different GDP indicators from the "nama_10_gdp" dataset
GDP <- getEurostatRCV(kod = "nama_10r_2gdp")
# For the "GDP.PC" dataframe, we only use EUR_HAB_EU	
# .. Euro per inhabitant in percentage of the EU average
GDP.PC <- GDP[GDP$unit == "EUR_HAB_EU",]
# ?cast
GDP.PC <- cast(GDP.PC, geo ~ time, value="value")
head(GDP.PC) # the data frame is organized so that 
#              it can be merged with the geodata. 
#              GDP.PC contains data for EU all countries at NUTS2
#              .. at this point, we do not filter for the 6 countries as in Step 1
```

## Step 3 - Merging spatial and macroeconomic data  

```{r}
require(broom)
colnames(GDP.PC) <- paste0("x",colnames(GDP.PC)) #its easier to work with column name "x2015" then with "2015", numeric column names can cause problems 
CE_map1 <- tidy(CE_map, "NUTS_ID") #will clean and transform spatialpolygondataframe to dataframe 
#each polygon is now set of points with latitude, logitude and with order in wich it should be drawn
#for example plot(CE_map1$long, CE_map1$lat)
CE_map1 <- merge(CE_map1, GDP.PC, by.x="id",by.y="xgeo")
#To ensure correct plotting of "hole" NUTS2 regions, we have to
# prepare a top layer to the plot. Relevant for: Bremen, Berlin, Praha, Wien.
CE_map.holes <- CE_map1[CE_map1$id %in% c("DE30","DE50","CZ01","AT13"),]
#tidy polylines for borders
CE_map2 <- tidy(CE_map.borders)
```

## Step 4 Plotting (GDP data for the year 2015)

```{r}
ggplot(data=CE_map1, aes(long, lat, group=group)) + 
  geom_polygon(aes(fill=x2015),color="grey50")+
  # Next line is the top layer for "hole" regions
  # .. if you comment-out, some "hole" regions may not display properly
  geom_polygon(data=CE_map.holes, aes(fill=x2015),color="grey50")+
  scale_fill_gradientn('% of EU \n average',
                       colours=brewer.pal(8, "Blues")) +
  geom_path(data=CE_map2) +
  coord_map()+
  #coord_map(project = "orthographic", xlim = c(-22, 34), ylim = c(35, 70))+
  ggtitle("GDP per capita, EUR, in % of EU average, year 2015") +
  theme_minimal()
```


## Step 5 - Combined choropleths (GDP data for the years 2005 & 2015) 

```{r}
# ggplot usualy prefers data in long format
# its easier to split/group graphs by variables, when they are stored as factors in one column
CE_map.long <- melt(data = CE_map1, id.vars = c("id", "long", "lat", "group"), 
                 measure.vars = c("x2005", "x2015"))
# 
levels(CE_map.long$variable) <- c("2005","2015")
#
CE_map.long.holes <- CE_map.long[CE_map.long$id %in% c("DE30","DE50","CZ01","AT13"),]
CE_map2 <- tidy(CE_map.borders)
```




```{r}
(gmap <- ggplot(data=CE_map.long, aes(long, lat, group=group)) + 
  geom_polygon(aes(fill=value),color="grey50")+
  # Next line is the top layer for "hole" regions
  # .. if you comment-out, some "hole" regions may not display properly
  geom_polygon(data=CE_map.long.holes, aes(fill=value),color="grey50")+
  scale_fill_gradientn('% of EU \n average',
                       colours=brewer.pal(8, "Blues"))+
  geom_path(data=CE_map2)+
  coord_map()+ggtitle("GDP per capita, EUR, in % of EU average, year 2015") +
  theme_minimal()+
  facet_grid(~variable)) # vertically aligned 
```


```{r}
gmap + facet_grid(variable~.)
```


## Step 6 Animated choropleths

```{r}
# where <- choose.dir("Select folder to save frames (png)?")
# dir.create(file.path(where,"img_gif"))
# 
# apply(CE_map1, 2, min) #finding the limits for pallete
# apply(CE_map1, 2, max)
# 
# sc <- scale_fill_gradientn('% of EU \n average',
#                        colours=brewer.pal(9, "Blues"), limits=c(15,240))
# 
# for(year in colnames(GDP.PC)[2:ncol(GDP.PC)]){
# png(file.path(where,paste0("img_gif/",year,".png")))
# print(ggplot(data=CE_map1, aes(long, lat, group=group)) + 
#   geom_polygon(aes_string(fill=year),color="grey50")+
#   # Next line is the top layer for "hole" regions
#   # .. if you comment-out, some "hole" regions may not display properly
#   geom_polygon(data=CE_map.holes, aes_string(fill=year),color="grey50")+
#   sc+
#   geom_path(data=CE_map2)+
#   coord_map()+ggtitle(paste("GDP per capita, EUR, in % of EU average, year",substr(year,2,5))) +
#   theme_minimal())
# dev.off()
# }

#http://gifmaker.org/
```


## Assigment 


```{r}
toc <- getEurostatTOC()
#View(toc)
dta <- getEurostatRCV(kod = "ei_lmhr_m")
dta <- dta[dta$unit=="PC_ACT" &
             dta$s_adj == "SA" &
              dta$indic == "LM-UN-T-TOT" & 
               dta$geo %in% c("CZ","SK","DE","PL") &
             dta$time == "2017M01", c("geo","value")]
require(dplyr)
dta <- getEurostatRCV(kod = "ei_lmhr_m") %>% filter(unit=="PC_ACT", 
                      s_adj=="SA", 
                      indic=="LM-UN-T-TOT", 
                      geo %in% c("CZ","SK","DE","PL", "AT","HU"),
                       time == "2018M01") %>% select(geo, value)

CE_map <- map[substr(map$NUTS_ID,1,2) %in% c("CZ","SK","DE","PL", "AT","HU") & map$STAT_LEVL_ == 0,]

CE_map1 <- tidy(CE_map, "NUTS_ID") 

CE_map1 <- merge(CE_map1, dta, by.x="id",by.y="geo")


ggplot(data=CE_map1, aes(long, lat, group=group)) + 
  geom_polygon(aes(fill=value),color="grey50")+
  scale_fill_gradientn('% of EU \n average',
                       colours=brewer.pal(8, "Blues")) +
  coord_map()+
  theme_minimal()
```

