---
title: "Missing data & Multiple Imputation (MI)"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("install_packages.r")
require(knitr)
require(ggplot2)
require(dplyr)
require(GGally)
require(lmtest)
```


## Example description

LRM:    $\textit{lwage} \sim \textit{(Intercept)} + \textit{educ} + \textit{tenure} + \textit{female} + \textit{south}$

A) Data:

    1) From "wage1.csv", we only retrieve a small data sample, n=50
       for further estimation

    2) In our n=50 sample, we simulate 6 and 4 wage observations in 2 regressors
        (10 missing values total).

B) Estimation & model comparison

    1) We estimate the benchmark LRM using "full" dataset of n=50
    2) We estimate the LRM using complete cases only (listwise deletion)
    3) We estimate the LRM using mean substitution
    4) Example of multiple imputation in R is provided


## A.1

```{r}
wageData <- read.csv("dta/wage1.csv")
```



```{r}
set.seed(300)
sample1 <- sample(nrow(wageData), size = 50, replace = F)
full <- wageData[sample1, ] 
```

This is the `n=50` sample for benchmark estimation
rownames in "full" dataframe correspond to randomly selected rows from "wageData"

```{r}
row.names(full) <- c(1:50) # fixes row names.
```

## A.2 Simulate wage missing data for two variables: educ and tenure



```{r}
set.seed(119)
sample2 <- sample(nrow(full), size = 20, replace = F) #original missing 4
sample3 <- sample(nrow(full), size = 20, replace = F) #original missing 6
wage <- full # step 1, copy all data
wage[sample2, "educ"] <-  NA  # step 2, generate NAs
wage[sample3, "tenure"] <-  NA
sum(complete.cases(wage)) # complete cases in the wage dataset
#fix(wage)
```

Due to missing data, we have lost a total of n = `r 50-(sum(complete.cases(wage)))` observations.

## B.1 Our benchmark model - with no missing data simulated (all n=50 observations used)

```{r}
LRM.bench <- lm(lwage~educ+tenure+female+south, data = full)
summary(LRM.bench)
```


## B.2 LRM on observations with missing data 

estimated on complete.cases only  rows with NA entries are excluded automatically
and compare coeffs and significances with "LRM.bench"

```{r}
LRM.cc <- lm(lwage~educ+tenure+female+south, data = wage)
summary(LRM.cc) 
```

## B.3 LRM on observations with missing data - mean substitution used

Mean substitution split to two steps for clarity, the following mean substitution routine can be performed more efficiently....

```{r}
wage.ms <- wage # We shall use a new data.frame to make the mean substitution
educ.mean <- mean(wage.ms$educ, na.rm=T )     # Calculate the means
tenure.mean <- mean(wage.ms$tenure, na.rm=T ) #
wage.ms[is.na(wage.ms$educ), "educ"] <- educ.mean         # Mean substitutions
wage.ms[is.na(wage.ms$tenure), "tenure"] <- tenure.mean     #
```


```{r}
LRM.ms <- lm(lwage~educ+tenure+female+south, data = wage.ms)
summary(LRM.ms)
```

Compare the three models: coefficients and VIFs

```{r}
require(lmtest)
coeftest(LRM.bench) # benchmark
coeftest(LRM.cc)    # missing data -> negative impact on statistical significance
coeftest(LRM.ms)    # Mean subistitution -> falsely "improved" results as compared to benchmark
```


```{r, echo=FALSE}
kable(data.frame(LRM.bench = coeftest(LRM.bench)[,1],
                pval = coeftest(LRM.bench)[,4],
                 LRM.cc = coeftest(LRM.cc)[,1],
                pval = coeftest(LRM.cc)[,4],
                 LRM.ms = coeftest(LRM.ms)[,1],
                pval = coeftest(LRM.ms)[,4]), 
      digits=3)
```

## B.4 Multiple imputation example - using the {mice} package
tested on version 2.25 of `mice`

```{r}
wage.mice <- wage[ , c(22,2,3,4,6,11)]
head(wage.mice,15)
```


We shall use a new data.frame for multiple imputation MI uses ML estimation and does NOT work if dataframe contains linearly dependent variables and/or combinations such as var1 & log(var1)..

Estimation using the `mice` package


```{r}
require("mice") # install.packages("mice")
#help(package=mice)
```

Create the imputation object

```{r}
#?mice
imputed.data <- mice(wage.mice, seed=200)
```

Imputation summary

see page 16 of the `mice` PDF file for "pmm" and other methods
pmm - Predictive mean matching

```{r}
imputed.data 
```

Actual imputed values for each of the 5 imputations

```{r}
imputed.data$imp$educ
imputed.data$imp$tenure
```

Estimation of a model using MI:

```{r}
#?with
LRM.mice <- with(imputed.data, lm(lwage~educ+tenure+female+south))
```

Estimation output

```{r}
LRM.mice
#?pool
pool(LRM.mice)
(LRM.MI <- summary(pool(LRM.mice)))
```

fmi 

  - fraction of missing information as defined in Rubin (1987)
  - Rubin (1987). Multiple Imputation for Nonresponse in Surveys.
  - John Wiley & Sons, New York.
      
lambda 

  - proportion of the total variance that is attributable to the missing data.
    
```{r, echo=FALSE}
kable(data.frame(LRM.bench = coeftest(LRM.bench)[,1],
                 LRM.cc = coeftest(LRM.cc)[,1],
                 LRM.ms = coeftest(LRM.ms)[,1],
                LRM.MI = LRM.MI[,1]),
      digits=3)
```



```{r}
## Cross Validation
cv.wage <- wageData[-sample1,]
cv.wage <- cv.wage[1:200,]

LRM.MI.prediction <- LRM.MI[1,1]+
  LRM.MI[2,1]*cv.wage$educ+
  LRM.MI[3,1]*cv.wage$tenure+
  LRM.MI[4,1]*cv.wage$female+
  LRM.MI[5,1]*cv.wage$south

mean((cv.wage$lwage-predict(LRM.bench, cv.wage))^2)
mean((cv.wage$lwage-predict(LRM.cc, cv.wage))^2)
mean((cv.wage$lwage-predict(LRM.ms, cv.wage))^2)
mean((cv.wage$lwage-LRM.MI.prediction)^2)
```






## Assignment 1

1) Open the "hp2_mi.csv" dataset.
   This dataset contains an amended HPRICE2 dataset (as used in Wooldridge)
   dataset is shorter and contains missing data (wage)

2) Use OLS to estimate the equation (complete cases only):

   $\textit{log(price)} <- \textit{(Intercept)} + \textit{log(nox)} + \textit{dist} + \textit{rooms} + \textit{stratio}$

where:
   
|            variable      |       description                      |
|--------------------------|----------------------------------------|
|            price         |       housing price, $                 |
|              nox         |  nitrox. concentr. in parts per   100m |
|              rooms       |    number of rooms                     |
|              dist        |      wght dist to 5 employ centers     |
|              stratio     |  average student-teacher ratio         |

   all observations are provided as average values for different districts/areas

3) Using `nrow()` and `sum(complete.cases())` functions, find out the
   proportion of rows with missing data in the dataset.

4) Perform MI, estimation and evaluation of the model based on MI
   hint: replicate the steps on lines 91-107
   Use `seed=200` argument for the `mice()` function


