---
title: "Non-gaussian responses: generalized linear mixed models"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("install_packages.r")
require(knitr)
require(ggplot2)
require(dplyr)
require(nlme)
require(lme4)
require(gridExtra)
require(VGAM)
require(MASS)
require(glmmML)
```
# Introduction

Data: Bangladesh Demographic and Health Survey (BDHS)

| Variable | Description                                                                                                                  |
|----------|------------------------------------------------------------------------------------------------------------------------------|
| comm     |  Community identifier                                                                                                        |
| wonid    |  Woman identifier                                                                                                            |
| antemed  |  Received antenatal care at least   once from a medically-trained provider, e.g. doctor, nurse or midwife  (1 = yes, 0 = no) |
| bord     |  Birth order of child (ranges from   1 to 13)                                                                                |
| mage     |  Mother’s age at the child’s birth   (in years)                                                                              |
| urban    |  Type of region of residence at   survey (1=urban, 0=rural)                                                                  |
| meduc    | Mother’s level of education at survey (1=none, 2=primary, 3=secondary or   higher)                                           |
| slam     |  Mother’s religion (1=Islam,   0=other)                                                                                      |
| wealth   | Household wealth index in quintiles (1=poorest to 5=richest)                                                                 |

 Our response variable is a binary indicator of whether a woman received antenatal care from a medically-trained provider (a doctor, nurse or midwife) at least once before her most recent live birth. 

At level 1, we consider variables such as a woman’s age at the time of the birth and education. Level 2 variables include an indicator of whether the region of residence is classified as urban or rural. 

$E(y_{ij}|u_i) = \beta_0 + u_i + \epsilon_{ij}, u_i \sim N(0, \sigma^2_u) \epsilon_i \sim N(0, \sigma^2)$

First fitted model with only intercept and comunity effect 

$\texit{logit}(p_{ij}) = \log\bigg(\frac{p_{ij}}{1-p_{ij}}\bigg) = \beta_0 + u_{0j}$

$\beta_0$ is for all comunities and $u_0j$ is for $j$. 


```{r}
BDHS <- read.table("dta/BDHS.txt", header = TRUE)
```

```{r}
fit <- glmer(antemed ~ (1 | comm), family = binomial("logit"), data = BDHS)
```

```{r}
summary(fit)
```


$\hat{\beta}_0 = 0.148$ Intercept for $j$ community is then $0.148 + u_{0j}$, variance of $u_{0j}$ is 1.464

The likelihood ratio statistic for testing the null hypothesis that variance of $u_{0j}$ is 0 can be done as two model comparision:
```{r}
fit0 <- glm(antemed ~ 1, family=binomial("logit"),data=BDHS)
test = -2*logLik(fit0) + 2*logLik(fit)
mean(pchisq(test,df=c(0,1),lower.tail=FALSE))
```

```{r}
library(lattice)
qqmath(ranef(fit, condVar=TRUE),strip=FALSE)$comm
```

```{r}
BDHS$magec <- BDHS$mage - mean(BDHS$mage)
fit2 <- glmer(antemed~magec+(1|comm),family=binomial("logit"),data=BDHS)
summary(fit2)
```


```{r}
predprob <- fitted(fit2)
```


```{r}
predlogit <- logit(predprob)
datapred=unique(data.frame(cbind(predlogit = predlogit, comm = BDHS$comm,mage = BDHS$mage)))
xyplot(predlogit ~ mage, data = datapred, groups = comm, type = "l", col = "blue", xlim = c(9, 51), ylim = c(-4, 4))

ggplot(datapred) + 
  geom_line(aes(mage,predlogit, group=comm))
```

For a woman of 22 years the log-odds of receiving antenatal care ranges from about −2.2 to 2.5 depending on which community she lives in. This translates to a range in probabilities of $\frac{\exp^{(−2.2)}}{1+\exp^{(−2.2)}}=0.10$ to e$\frac{\exp^{(−2.5)}}{1+\exp^{(−2.5)}}=0.92$, so there are strong community effects.

```{r}
BDHS$meduc <- factor(BDHS$meduc)
BDHS$wealthc <- BDHS$wealth-mean(BDHS$wealth) # we center wealth
fit3 <- glmer(antemed~magec+wealthc+meduc+(1|comm), family=binomial("logit"), data=BDHS)
summary(fit3)
```


** Variance Partition Coefficient**

$$  \frac{\sigma^2_u}{\sigma^2_u + \sigma^2} = \frac{0.868}{0.868+3.29}=0.208$$

```{r}
as.numeric(summary(fit3)$varcor)
```

** Predict probabilities from a Multilevel model **

$$p_{ij} = \frac{e^{z_{ij}}}{1+e^{z_{ij}} }$$
where 

$$ z_{ij} = -0.491 - 0.005 \textit{magec}_{ij} + 0.403 \textit{welthc}_{ij} + 0.545 \textit{meduc2}_{ij}+  1.306 \textit{meduc3}_{ij}$$
**Two-level random slope model**

$$ log\bigg(\frac{p_{ij}}{1-p_{ij}}\bigg) = \beta_0 + \beta_1 \textit{magec}_{ij} + \beta_3 \textit{meduc}_{ij} + \beta_4 \textit{welthc}_{ij} + \beta_5 \textit{wealthc}_{ij} + u_{0j} + u_{4j} \textit{wealthc}_{ij}$$
the $\beta_{4j} = \beta_4 + u_{4j}$

```{r}
fit4 <- glmer(antemed~magec+meduc+wealthc+(1+wealthc|comm), data = BDHS, family=binomial("logit"))
summary(fit4)
```

```{r}
anova(fit3,fit4)
```

$$ log\bigg(\frac{p_{ij}}{1-p_{ij}}\bigg) = (-0.4908 + \hat{u}_i) + (0.408 + \hat{v}_i) \textit{wealthc}_{ij}$$

We can include a Level 2 explanatory variable `urban`
```{r}
BDHS$urban <- factor(BDHS$urban)
levels(BDHS$urban) <- c("rural","urban")
fit5 <- glmer(antemed~magec+meduc+wealthc+urban+(1+wealthc|comm), 
              data = BDHS, family=binomial("logit"))
summary(fit5)
```








### Deer Data Assigment 

Data from Zurr - Mixed Effects Models and Extensions in Ecology with R

Relationship between the presence and absence of larvae (Elaphostrongylus cervi) in deer across Spain and the explanatory variables `length` and  `sex` of the deer and the `Farm`.




```{r}
deer <- read.table('dta/deer.txt',header=TRUE)
deer$Farm <- factor(deer$Farm)
ggplot(deer) + geom_point(aes(Length, as.factor(infect)),size=4, alpha=1/10)
ggplot(deer) + geom_boxplot(aes(as.factor(infect), Length))
ggplot(deer) + geom_density(aes(Length, group=as.factor(infect), fill= as.factor(infect)),alpha=1/5)+
   theme_bw()
```


**Estimate model with fixed effect**

```{r}

```


**Estiamte model with random effec**

$$ y_{ij} \sim Bin(1, p_{ij})$$
$$\textit{logit}(p_{ij}) = \beta_0 + \beta_1 \textit{Length}_{ij} + u_i, \quad u_i \sim N(0, \sigma_u^2)$$
where $y_{ij}$ is 1 if deer $j$ from Famr $i$ is infected. 


```{r}
deer.glmm <- glmer(infect~Length+(1|Farm),data=deer,family=binomial)

# model without the farm
deer.glm <- glm(infect~Length,data=deer,family=binomial)
```



```{r, warning=FALSE}
ggplot(deer) + 
  geom_point(aes(Length, infect, color=Farm),
             size=4,
             alpha=9/10) + 
  geom_smooth(aes(Length,fitted(deer.glmm), group=Farm, color=Farm),
              method="loess")+
  geom_smooth(aes(Length,fitted(deer.glm)),
              size=2, 
              color="black",
              method="lm",
              formula = y~poly(x,4), 
              se=FALSE)+
  theme_bw()
```





This example was based on lecture by [Dae-Jin Lee](http://idaejin.github.io/bcam-courses/neiker-2016/material/mixed-models/) and slightly modified. [cc](https://creativecommons.org/licenses/by-sa/3.0/)