---
title: "Multilevel models"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("install_packages.r")
require(knitr)
require(ggplot2)
require(dplyr)
require(nlme)
require(lme4)
```

### Introduction 

High School & Beyond [^1](https://en.wikipedia.org/wiki/High_School_and_Beyond]) is a nationally representative survey of U.S. public and Catholic high schools conducted by the National Center for Education Statistics (NCES). The data are a subsample of the 1982 HSB survey with 7,185 students from 160 schools. The average sample size per school is approximately 45 students.


This example was based on lecture by [Dae-Jin Lee](http://idaejin.github.io/bcam-courses/neiker-2016/material/mixed-models/) and slightly modified. [cc](https://creativecommons.org/licenses/by-sa/3.0/)

| Variable | Description                                                          |
|----------|----------------------------------------------------------------------|
| School   |  an   ordered factor identifying the school that the studend attends |
| Sex      |  Female or Male                                                      |
| SES      |  a standardized scale of   socio-economic status                     |
| MathAch  |  a measure of mathematics   achievement                              |
| CSES     |  centered mean of the SES values   for the school                    |


```{r}
MathAchieve <- readRDS("dta/MathAchieve.Rds")
names(MathAchieve)
summary(MathAchieve)
```

How to center data  
```{r, eval=FALSE}
MathAchieve %>% group_by(School) %>% mutate(CSES = floor(100*(SES-mean(SES)))/100) %>% ungroup %>% select(CSES) 
MathAchieve$CSES[1:10]
```

$$y_j = \beta_0 + \beta_1 x_j + \epsilon_j$$
$$ \textit{MathAch}_j = \beta_0 + \beta_1 \textit{CSES}_j + \epsilon_j$$ 
This model ignores that the students attends different schools ($j^{th}$ subindex corresponds to the individual unit level).


```{r}
(multilev0 <- lm(MathAch~CSES, data=MathAchieve))
```


```{r, echo=FALSE}
plot(MathAchieve$CSES,MathAchieve$MathAch,cex=.5,col="grey")
abline(multilev0,col=2,lwd=3)
```


Suppose we are interested in comparing differences between schools.


$$y_j = \beta_0 + \beta_1 x_{ij} + \epsilon_j$$
where the subindex $i$ indicates the school of the student, with $\beta_0$ we specify a separate intercept for each school. Indeed, what we include is a categorical variable with as many categories as schools. Model multilev1 considers the Schools as a fixed effect, i.e. we implicitly assume that we are interested in the schools of the survey.


```{r, eval=FALSE}
summary(lm(formula = MathAch ~ CSES + School, data = MathAchieve))
```

We can complicate more the model adding a different socio-economic status for each school, i.e. 1 unit increase of socio-econonomic status can explain different maths achievements in different schools. A model like:
$$y_{ij}=\beta_{0i} + \beta_{1i}x_{x_{ij}} + \epsilon_{ij}$$
now we have the subindex $i$ for the slope $\beta_{1i}$.

We are not interested in these schools in particular, but in the population of schools in order to compare schools with different characteristics.


## Multilevel model for group means

Hierarchical data are collected when sampling takes place at two or
more levels, one nested within the other. Some examples:
  - Students within schools (two levels).
  - Students within classrooms within schools (three levels).



```{r}
ggplot(MathAchieve[School== c("91","3","31","52","74"),])+geom_boxplot(aes(School, MathAch))
ggplot(MathAchieve[School== c("91","3","31","52","74"),])+geom_point(aes(School, MathAch))
```


We specify two levels in the model:
- Level 1: $y_{ij} = \mu_i + \epsilon_{ij}$
- Level 2: $\mu_i = \beta_0 + u_i$
where $\mu_i$ is average of each school,$\beta_0$ is average of all students, $u_i$ is deviation of $i$th school from total average. Then we can write previous model as: 

$$ y_{ij} = \beta_0 + u_i + \epsilon_{ij}, \quad i=1,\dots,m \quad j=1,\dots,n_m$$
At both levels, we assume $u_i\sim N(0,\sigma^2_{u_i})$ and $\epsilon_{ij} \sim N(0,\sigma^2)$ (both independent).

The mean of $y$ for the $i$th group is given by $\beta_0 + u_i$ , then $u_i$ is the deviation of the mean of iith group to the total average. Individual level residuals $\epsilon_{ij}$ are the difference between the value of the response variable of the $j$jth individual and the average of the group they belong. 

We fit the model with `lme`

```{r}
multilev.lme <- lme(MathAch~1,random=~1|School, data=MathAchieve)
multilev.lme
```


```{r}
VarCorr(multilev.lme)
```



get beta's and predict

```{r}
(beta0 <-  fixef(multilev.lme))
```



```{r}
fixef(multilev.lme)-(predict(multilev.lme,
                             newdat=list(School=c("91","3","31","52","74"))))
```

```{r}
multilev.lmer<-lmer(MathAch~1+(1|School),data=MathAchieve)
rr1 <- ranef(multilev.lmer, condVar = TRUE)
rr1 <- data.frame(rr1) %>% mutate(sd1 = condval + condsd, sd2 = condval- condsd) 
ggplot(rr1, aes(as.numeric(grp),condval))+
  geom_ribbon(aes(ymin=sd2, ymax=sd1), fill = "lightblue")+
    geom_point()+
  xlab("schools")+
  ylab("intercept")+
  theme_minimal()
```

If we had considered a fixed effect model, we would get a single variance $\sigma^2=6.882$, $\sigma^2=6.882$ obtained from `summary(multilev.NULL)$sigma` Considering a random effects model, we have σ2uσu2 and σ2σ2, and the total variability of the data would be given by:
$$Var(y_{ij}) = \sigma^2 + \sigma^2_u$$
i.e. the variability within the schools plus the variability among the schools. The intra-class correlation coefficient would be
$$ICC=\frac{2,93^2}{2,93^2+6,27^2}=0,18$$
Then the 18% of the total variability is due to the difference among the averages of the schools.


## Random slopes models


```{r}
(ml3 <- lme(MathAch ~ CSES, random = ~CSES|School, data=MathAchieve))
```


