---
title: "Poisson regression"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("install_packages.r")
require(knitr)
require(ggplot2)
require(dplyr)
```

### Introduction 

Case study: Lung cancer incidence in Denmark 1968{1971

This data set contains counts of incident lung cancer cases and population size in
four neighbouring Danish cities by age group.

| Variable | Description                                                     |
|----------|-----------------------------------------------------------------|
| city     |  factor with levels Fredericia,   Horsens, Kolding, and Vejl    |
| age      |  factor with levels 40-54, 55-59,   60-64, 65-69, 70-74, and 75 |
| pop      |  the number of inhabitants.                                     |
| cases    |   the number of lung cancer cases.                              |


### Loading data 

```{r}
lung <- read.table("dta/lung.txt",header=TRUE)
```


```{r}
head(lung)
```

Incident rate is $\lambda = \frac{\textit{cases}}{\textit{pop}}$. In Fredericia for people aged 40-45 it is `r lung %>% filter(city=="Fredericia"&age=="40-54") %>% mutate(lambda = cases/pop) %>% select(lambda)`. 


```{r, echo=FALSE}
boxplot(cases~age,data=lung,col="bisque",main="How does the expected number of lung cancer counts vary by age?")
```

### Poisson regression


We start considering a model with age as covariate
$$\log(\lambda_i) = \beta_0 + \beta_1 I(\textit{Age55-59}_i) + \beta_2 I(\textit{Age60-64}_i) + \beta_3 I(\textit{Age65-69}_i) + \beta_4 I(\textit{Age70-74}_i)  + \beta_5 I(\textit{Age>74}_i) $$

where $I(.)$ is a indicator (1 if TRUE, 0 otherwise) for each range of age, with
Age40-45 is used as baseline.



```{r}
lungmod1 <- glm(cases ~ age, family=poisson, data=lung)
summary(lungmod1)
```


**Interpretation**:

- `exp(2.11021) = 8.24` is the expected count of cancer cases among individuals
aged 40 - 54
- `exp(2.1021 - 0.03077) = 8.00` is the expected count of cancer cases among
individuals aged 55 - 59
- `exp(-0.0377) = 0.97` is the ratio of the expected counts comparing the
55 - 59 aged group to the baseline group of age `40 - 54. exp(`$\hat{\beta}_1$`)` is also the
relative rate.


If we calculate the CI's for all ages we find that all contain 0, is there any association
between cancer and age?

```{r}
confint(lungmod1)
```



```{r}
anova(lungmod1,test="Chisq")
```




### Likelihood Ratio Test

$H_0: \beta_1=\beta_2=\dots=\beta_i$

$H_1: non H_0$

```{r}
lungmod0 <- glm(cases ~ 1, family=poisson, data=lung)
anova(lungmod0,lungmod1,test="Chisq")
```

and hence, we do not reject the hyphotesis of all $\beta_i = 0$, for $i = 1,2,\dots,5$.

### How to improve the analysis?


- We have considered the counts of lung cancer cases.
- Each city and age group has a different population size.


```{r, echo=FALSE}
par(mfrow=c(1,2))
hist(lung$pop,col="lightgrey", main="Population size")
barplot(xtabs(pop~age,data=lung),main="Population counts by age group")
```


- So far, we have modeled expected counts for each population group, within the **4**
year period of time, i.e., intrinsically `rate = counts/'4 years'`
- It may be of more interest to know the rate per person, per 4 period of observation.
We are interested in $r_i = \frac{\lambda_i}{\textit{pop}_i} = \frac{E(\textit{count}_i)}{\textit{pop}_i}$ and model it by a log-linear model
- Then, our model is $Y_i \sim \textit{Pois}(\lambda_i) = \textit{Pois}(r_i \times \textit{pop}_i)$
- On a log-scale, our model is: 

$$\log(\frac{\lambda_i}{\textit{pop}_i}) = \beta_0 + \beta_1 I(\textit{Age55-59}_i) + \beta_2 I(\textit{Age60-64}_i) + \beta_3 I(\textit{Age65-69}_i) + \beta_4 I(\textit{Age70-74}_i)  + \beta_5 I(\textit{Age>74}_i)$$


- All counts are restricted to the same period of 1968 - 1971, the value  $\lambda_i$ are rates `per
4-years'
- To obtain an easier interpretation of the rates we can divide by 4 to get rate per person-year
and multiply by 10.000 to get a rate per 10.000 person-years, i.e. divide by 2500

This can be easily done by

```{r}
lungmod2 <- glm(cases ~ age + offset(log(pop/2500)), 
                family=poisson, data=lung)

lungmod4 <- glm(cases ~ city + offset(log(pop/2500)), 
                family=poisson, data=lung)


summary(lungmod2)
```


- `log(pop/2500)` is the `offset` (try `?offset`)
- The **offset** accounts for the population size, which could vary by age, region, etc ...
- It gives a convenient way to model rates per person-years, instead of modeling the raw
counts.



Confidence intervals for the parameters can also be obtained as:

```{r}
confint(lungmod2)
```

The inclusion of the offset, implies that the interpretation of the coefficients
should be done in terms of $\log(\lambda_i) - \textit{offset_i}$
In our example, $\lambda_i$ is the expected number of cases observed in a particular
age group, within a 4 year period of time.
Hence in our case, with an offset of `log(pop/2500)`, we should think of the
outcome as log rate per 10,000 person years.

**Interpretation**

- $\beta_0$ is the log rate of cancer cases per 10.000 person years in the age group of
40 - 54 (baseline)
- $\beta_1$ is the log relative rate of cancer cases per 10.000 person years comparing
the age group of 50 - 59 to the baseline age group 40 - 5
- $\beta_2$ is the log relative rate of cancer cases per 10.000 person years comparing
the age group of 60 - 64 to the baseline age group 40 - 54


Note that, including the offset, the regression coeffcients are signifficant in the
expected counts per year at age group compared to the baseline group.

Now, with the offset, we are looking for differences in the expected counts per
person-year, across age groups.

### Likelihood Ratio Test

```{r}
lungmod3<-glm(cases ~ 1, family = poisson, data = lung, 
              offset = (log(pop/2500)))
anova(lungmod3,lungmod2,test="Chisq")
```

| Without offset                       | With offset                                           |
|--------------------------------------|-------------------------------------------------------|
| Do not reject $H_0$                  | Reject $H_0$                                          |
| Expected cases across the age groups | Expected cases per person year, across the age groups |


There is no difference because the lower population Age75+ might be
counterbalanced by high rate of cases with increasing age

The offset let us compare the rate of cancer among those who are alive, i.e. taking
into account the number of people within cohort of interest.


cross-classifying table


```{r}
xtabs(pop~age,data=lung)
xtabs(cases~age,data=lung)
```

Relative rates

```{r}
exp(coef(lungmod2))
```


### Assigment 

Use Poisson regression to model the number of Insurace claims during a given period. Use Group /engine capacity/ and car Age as regressors. For offset, use the number of insurance-holders per Group.

The data given in data frame Insurance consist of the numbers of policyholders of an insurance company who were exposed to risk, and the numbers of car insurance claims made by those policyholders in the third quarter of 1973.

| Variable | Description                                                                                 |
|----------|---------------------------------------------------------------------------------------------|
| District | factor: district of residence of   policyholder (1 to 4): 4 is major cities.                |
| Group    | an ordered factor: group of car   with levels <1 litre, 1–1.5 litre, 1.5–2 litre, >2 litre. |
| Age      | an ordered factor: the age of the   insured in 4 groups labelled <25, 25–29, 30–35, >35.    |
| Holders  | numbers of policyholders.                                                                   |
| Claims   | numbers of claims                                                                           |



```
Insurance <- read.csv("dta/Insurance.csv") ....
head(Insurance)
str(Insurance)
#
# Estimate the Poisson regression model
#
# Produce Confidence intervals and comment on regressor significance.
```


