---
title: "Quantile Regression"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(foreign)
require(knitr)
require(quantreg)
require(ggplot2)
require(gridExtra)
require(RColorBrewer)
require(splines)
```

### Introduction 

example from MICROECONOMETRICS USING STATA - A. Colin Cameron and Pravin K. Trivedi 
package required for quantile regression is `guantreg`


### Loading data 

for loading Stat data format we can use `read.dta` from `foreign`

```{r load_data}
dta <- read.dta("dta/mus03data.dta")
apply(dta, 2,function(x) sum(is.na(x))) # number of NA in each column
```


```{r}
dta <- dta[,c("ltotexp", "suppins","totchr", "age", "female", "white")]
dta <- na.omit(dta)
```


| Variable | Description                  | Type    |
|----------|------------------------------|---------|
| ltotexp  | log of medical expenditure   | num     |
| suppins  | supplement private insurance | binary  |
| totchr   | healt status                 | num     |
| age      | age 65-90                    | integer |
| female   | gender                       | binary  |
| white    | race                         | binary  |



### Descriptive statistics

```{r}
summary(dta)
```


### Plots

```{r, echo=FALSE}
ggplot(dta) + 
  geom_boxplot(aes(as.factor(age), ltotexp)) + 
  facet_grid(~white) + xlab("Age") + ggtitle("Boxplot of ltotexp by age and race")
#--------------------------------
ggplot(dta) + 
  geom_boxplot(aes(as.factor(suppins), ltotexp)) + 
  facet_grid(~white) + xlab("Supplement private insurance") + ggtitle("Boxplot of ltotexp by supplement private insurance and race")
#--------------------------------
ggplot(dta) + 
  geom_boxplot(aes(as.factor(totchr), ltotexp)) + 
  facet_grid(~white) + xlab("Health status") + ggtitle("Boxplot of ltotexp by health status and race")
#--------------------------------
colourCount = length(unique(dta$age))
getPalette = colorRampPalette(brewer.pal(9, "YlGnBu"))
ggplot(dta) + geom_histogram(aes(totchr, fill=as.factor(age), group= age), bins=7) + 
  scale_fill_manual(values = getPalette(colourCount))+ggtitle("Histogram of totchr")+guides(fill=guide_legend(title="Age"))

```


```{r , echo=FALSE}
(quantile(dta$ltotexp, c(0.1, 0.5, 0.9)))
plot(seq(0,1, by = (1/(2955-1))),dta$ltotexp, type="s", main = "ln(totexp)",xlab="fraction of the data", ylab="quantiles of ln(totexp) if totexp >0")
abline(v=c(0.1, 0.5, 0.9), col="red")
```

#OLS regression 

```{r}
summary(lm.model <- lm(ltotexp~suppins+age+female+white+totchr, data=dta))
```

```{r, echo=FALSE}
resid <- residuals(lm.model)
fitted <- fitted(lm.model)
grid.arrange(
ggplot()+geom_point(aes(fitted,resid))+geom_hline(yintercept=mean(resid), colour="red"),
ggplot()+geom_boxplot(aes(as.factor(dta$age),resid))+geom_hline(yintercept=mean(resid), colour="red"),
ggplot()+geom_boxplot(aes(as.factor(dta$totchr),resid))+geom_hline(yintercept=mean(resid), colour="red"),
ggplot()+geom_line(aes(1:length(dta$ltotexp),dta$ltotexp))+geom_line(aes(1:length(dta$ltotexp),fitted, colour="red")),
  ncol = 2, nrow = 2)
```


### Median regression

```{r}
summary(qr.model.5 <- rq(ltotexp~., data=dta, tau=0.5))
```


```{r}
#effects(qr.model.5)
```
### Comparing Qunatile Regression and OLS

```{r , echo=FALSE}
df.compare <- data.frame(OLS.coef = summary(lm.model)$coefficients[,1], OLS.tval= summary(lm.model)$coefficients[,3])
df.compare.coef <- data.frame(OLS.coef = summary(lm.model)$coefficients[,1])
df.compare.tval <- data.frame(OLS.tval= summary(lm.model)$coefficients[,3])
for(q in c(0.10, 0.25, 0.50, 0.75, 0.90)){
  model <- rq(ltotexp~., data=dta, tau=q)
  assign(paste0("qr.model",q), model)
  df.compare<-cbind(df.compare, summary(model)$coefficients[,1], summary(model)$coefficients[,3])
  df.compare.coef<-cbind(df.compare.coef, summary(model)$coefficients[,1])
  df.compare.tval<-cbind(df.compare.tval,summary(model)$coefficients[,3])
}

colnames(df.compare) <- c("OLS.coef","OLS.tval", "QR.10.coef", "QR.10.tval", "QR.25.coef", "QR.25.tval", "QR.50.coef", "QR.50.tval", "QR.75.coef", "QR.75.tval", "QR.90.coef", "QR.90.tval")
colnames(df.compare.coef) <- c("OLS.coef", "QR.10.coef", "QR.25.coef",  "QR.50.coef", "QR.75.coef",  "QR.90.coef")
colnames(df.compare.tval) <- c("OLS.tval", "QR.10.tval",  "QR.25.tval",  "QR.50.tval",  "QR.75.tval",  "QR.90.tval")
```


```{r, echo=FALSE}
kable(df.compare)
kable(df.compare.coef)
kable(df.compare.tval)
#df.compare
#df.compare.coef
#df.compare.tval
```


```{r}
anova(qr.model0.25, qr.model0.5)
anova(qr.model0.5, qr.model0.75)
anova(qr.model0.25,qr.model0.75)
```

$H_0$ no difference in coefficients


```{r, echo=FALSE}
plot(qr.model<-rq(ltotexp~., data=dta, tau=seq(.1,0.9,0.1)), mar = c(5.1, 4.1, 2.1, 2.1), xlab="Quantile", mfrow=c(2,3) )
```


```{r, echo=FALSE}
qr.model_plot<-rq(ltotexp~totchr, data=dta, tau=c(0.1,0.25,0.5,0.75,0.09))
resid <- residuals(qr.model_plot)
fitted <- fitted(qr.model_plot)
ggplot()+ 
  geom_line(aes(dta$totchr,fitted[,1], color=paste("tau 0.1")))+
  geom_line(aes(dta$totchr,fitted[,2], color=paste("tau 0.25")))+
  geom_line(aes(dta$totchr,fitted[,3], color=paste("tau 0.5")))+
  geom_line(aes(dta$totchr,fitted[,4], color=paste("tau 0.75")))+
  geom_line(aes(dta$totchr,fitted[,5], color=paste("tau 0.90")))+
  xlab("totchr")+ylab("fitted values")
```


