---
title: "Treatment effects - Catholics"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
source("install_packages.r")
require(knitr)
require(dplyr)
require(plotly)
require(ggplot2)
require(gridExtra)
require(MatchIt)
require(ATE)
```



## Data

```{r}
dta <- read.csv("dta/ecls.csv")
```

Early Childhood Longitudinal Study [United States]: Kindergarten Class of 1998-1999, Third Grade (ICPSR 4075) [here](http://www.researchconnections.org/childcare/studies/4075?q=c5r2mtsc&type=Data+Sets]). Original analysis [here](https://sejdemyr.github.io/r-tutorials/statistics/tutorial8.html)


| childid       | child identification number                                                                       |                          |
|---------------|---------------------------------------------------------------------------------------------------|--------------------------|
| catholic      | dummy: 1= catholic                                                                                | 0 = public               |
| race          | child composite race                                                                              |                          |
| race_white    | dummy: 1= race==white                                                                             | 0= not white             |
| race_black    | dummy: 1= race==black                                                                             | 0= not black             |
| race_hispanic | dummy: 1= race==hispanic                                                                          | 0= not hispanic          |
| race_asian    | dummy: 1= race==asian                                                                             | 0= not asian             |
| p5numpla      | fall 3rd grade: # of places child has lived for at least 4 months since 1st grade                 |                          |
| p5hmage       | fall 3rd grade: age of mother                                                                     |                          |
| p5hdage       | fall 3rd grade: age of father                                                                     |                          |
| w3daded       | fall 3rd grade: father's education level                                                          |                          |
| w3momed       | fall 3rd grade: mother's education level                                                          |                          |
| w3daded_hsb   | dummy: 1= father's education level is high school or below                                        | 0= some college or above |
| w3momed_hsb   | dummy: 1= mother's education level is high school or below                                        | 0= some college or above |
| w3momscr      | fall 3rd grade: mother's occupational prestige score                                              |                          |
| w3dadscr      | fall 3rd grade: father's occupational prestige score                                              |                          |
| w3inccat      | fall 3rd grade: family income category (13 categories)                                            |                          |
| w3income      | fall 3rd grade: numeric family income (recoded by taking midvalue from each category of w3inccat) |                          |
| w3povrty      | poverty score dummy                                                                               |                          |
| p5fstamp      | food stamp dummy                                                                                  |                          |
| c5r2mtsc      | 3rd grade math t-score                                                                            |                          |
| c5r2mtsc_std  | standardized values of c5r2mtsc                                                                   |                          |

## Data exploration 

```{r}
summary(dta)
sum(dta$catholic)
sum(dta$catholic)/nrow(dta)
colnames(dta)
apply(dta, 2, function(x) sum(is.na(x)))
```

```{r, echo=FALSE}
grid.arrange(
  ggplot(dta) + geom_boxplot(aes(as.factor(catholic),c5r2mtsc_std)),
  ggplot(dta) + geom_boxplot(aes(as.factor(catholic),w3income)),
  ncol=2
  )

grid.arrange(
  ggplot(dta[dta$catholic==0,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(race_black))) + 
    coord_polar(theta = "y") + 
    guides(fill=guide_legend(title="BLACK")),
  ggplot(dta[dta$catholic==1,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(race_black))) +
    coord_polar(theta = "y") +
    guides(fill=guide_legend(title="BLACK")),
  ncol=2
)
grid.arrange(
  ggplot(dta[dta$catholic==0,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(race_hispanic))) + 
    coord_polar(theta = "y") + 
    guides(fill=guide_legend(title="HISP"))+ylab("untreated"),
  ggplot(dta[dta$catholic==1,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(race_hispanic))) +
    coord_polar(theta = "y") +
    guides(fill=guide_legend(title="HISP"))+ylab("treated"),
  ncol=2
)
```



```{r, echo=FALSE}
g <- ggplot(dta) + xlab("Treatment (1,0)")
grid.arrange(
  g + geom_boxplot(aes(as.factor(catholic),w3income)),
  g + geom_boxplot(aes(as.factor(catholic),w3dadscr)),
  g + geom_boxplot(aes(as.factor(catholic),w3momscr)),
  ncol=3
  )
```

# Pre-analysis using non-matched data


```{r}
dta %>%
  group_by(catholic) %>%
  summarise(n_students = n(),
            mean_math = mean(c5r2mtsc_std),
            std_error = sd(c5r2mtsc_std) / sqrt(n_students))
```

This is how the math score is standardized
```{r}
dta %>%
  mutate(test = (c5r2mtsc - mean(c5r2mtsc)) / sd(c5r2mtsc)) %>% 
  group_by(catholic) %>%
  summarise(mean_math = mean(test))
```



```{r}
t.test(dta$c5r2mtsc_std~dta$catholic)
#or with with()
#with(ecls, t.test(c5r2mtsc_std ~ catholic))
```


## Difference in means: pre treatment cov

Selected variables 
```{r}
ecls_cov <- c('race_white', 'p5hmage', 'w3income', 'p5numpla', 'w3momed_hsb')
```


```{r}
dta %>% group_by(catholic) %>% select(one_of(ecls_cov)) %>% summarise_all(funs(mean(.)))
```
```{r}
lapply(ecls_cov, function(x){t.test(dta[,x]~dta[,"catholic"])})
```

## Propensity score estimation


```{r}
dta <- dta %>% mutate(w3income_1k = w3income/1000)
ecls_cov <- c('race_white', 'p5hmage', 'w3income_1k', 'p5numpla', 'w3momed_hsb')
formulaglm1 <- paste0("catholic~",paste0(ecls_cov,collapse="+"))
summary(glm1 <- glm(formulaglm1,   family = binomial(), data=dta))
```

```{r}
prs_df <- data.frame(pr_score = predict(glm1, type = "response"),
                     catholic = glm1$model$catholic)
sample_n(prs_df, size=5)
```

```{r, echo=FALSE}
grid.arrange(
  ggplot(prs_df[prs_df$catholic==0,])+
    geom_histogram(aes(x=pr_score))+ggtitle("")+
    xlab("Probability of going to Catholic school"),
  ggplot(prs_df[prs_df$catholic==1,])+
    geom_histogram(aes(x=pr_score))+
    xlab("Probability of going to Catholic school"),
  ggplot(prs_df)+
    geom_density(aes(x=pr_score, 
                     group=as.factor(catholic),
                     color=as.factor(catholic), 
                     fill=as.factor(catholic)),
                 alpha = 0.2)+
    xlab("Probability of going to Catholic school")
)
```

## Matching algorithm 

MatchIt does not allow missing values
```{r}
dta <- dta %>%  
  select(c5r2mtsc_std, catholic, one_of(ecls_cov)) %>%
  na.omit()
```


```{r}
mod_match <- matchit(catholic~race_white+p5hmage+w3income_1k+p5numpla+w3momed_hsb, method="nearest",distance="logit", data=dta)
```


```{r}
dta.m <- match.data(mod_match)
dim(dta.m)
colnames(dta.m)
```


```{r, echo=FALSE}
fn_bal <- function(dta, variable) {
  dta$variable <- dta[, variable]
  support <- c(min(dta$variable), max(dta$variable))
  ggplot(dta, aes(x = distance, y = variable, color = as.factor(T))) +
    geom_point(alpha = 0.2, size = 1.3) +
    geom_smooth(method = "loess", se = F) +
    xlab("Propensity score") +
    ylab(variable) +
    theme_bw() +
    ylim(support)
}
grid.arrange(
fn_bal(dta.m, "race_white"),
fn_bal(dta.m, "p5hmage"),
fn_bal(dta.m, "w3income_1k"),
fn_bal(dta.m, "p5numpla"),
fn_bal(dta.m, "w3momed_hsb"),
nrow=4)

```

```{r}
dta.m %>% group_by(catholic) %>% select(ecls_cov) %>% summarise_all(funs(mean(.)))
```

```{r}
lapply(ecls_cov, function(x){t.test(dta.m[,x]~dta.m[,"catholic"])})
```

```{r}
t.test(dta.m$c5r2mtsc_std~dta.m$catholic)
```


```{r, echo=FALSE}
grid.arrange(
  ggplot(dta.m) + geom_boxplot(aes(as.factor(catholic),c5r2mtsc_std)),
  ggplot(dta.m) + geom_boxplot(aes(as.factor(catholic),w3income_1k)),
  ncol=2
  )

grid.arrange(
  ggplot(dta.m[dta.m$catholic==0,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(race_white))) + 
    coord_polar(theta = "y") + 
    guides(fill=guide_legend(title="WHITE")),
  ggplot(dta.m[dta.m$catholic==1,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(race_white))) +
    coord_polar(theta = "y") +
    guides(fill=guide_legend(title="WHITE")),
  ncol=2
)

prs_df2 <- data.frame(pr_score = predict(glm(formulaglm1,   family = binomial(), data=dta.m), type = "response"),
                     catholic = dta.m$catholic)
  ggplot(prs_df2)+
  geom_density(aes(x=pr_score, 
                     group=as.factor(catholic),
                     color=as.factor(catholic), 
                     fill=as.factor(catholic)),
                 alpha = 0.2)+
    xlab("Probability of going to Catholic school")
```


```{r}
lm_treat1 <- lm(c5r2mtsc_std ~ catholic, data = dta.m)
summary(lm_treat1)
```

```{r}
lm_treat2 <- lm(c5r2mtsc_std ~ catholic + race_white + p5hmage +
                 w3income_1k+ p5numpla + w3momed_hsb, data = dta.m)
summary(lm_treat2)
```


## Package ATE

```{r}
Y<-dta[,"c5r2mtsc_std"]
treat<-dta[,"catholic"]
X<-dta[,colnames(dta)%in%ecls_cov]

fit1<- ATE(Y,treat,X)
fit1
summary(fit1)
plot(fit1)
```


 
 