---
title: "Treatment effects"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
source("install_packages.r")
require(knitr)
require(dplyr)
require(plotly)
require(ggplot2)
require(gridExtra)
require(MatchIt)
require(ATE)
```



## Data

```{r}
dta <- read.csv("dta/TableF19-3.csv")
```

LaLonde (1986) Earnings Data, 2,490 Control Observations and 185 Treatment Observations 
Source: LaLonde (1986).

| Variable | Description                              |   |
|----------|------------------------------------------|---|
| t        | Treatment dummy variable,                |   |
| age      | Age in years,                            |   |
| educ     | Education in years,                      |   |
| marr     | Dummy variable for married,              |   |
| black    | Dummy variable for black,                |   |
| hisp     | Dummy variable for Hispanic,             |   |
| nodegree | Dummy for no degree (not used),          |   |
| re74     | Real earnings in 1974, scaled by 10,000, |   |
| re75     | Real earnings in 1975, scaled by 10,000, |   |
| re78     | Real earnings in 1978, scaled by 10,000, |   |
| age2     | Age squared,                             |   |
| educ2    | Educ squared,                            |   |
| re742    | Re74 squared                             |   |
| re752    | Re75 squared                             |   |
| blacku74 | Black times 1(re74                       |   |

## Data exploration 

```{r}
summary(dta)
sum(dta$T)
```

```{r, echo=FALSE}
grid.arrange(
  ggplot(dta) + geom_boxplot(aes(as.factor(T),AGE)),
  ggplot(dta) + geom_boxplot(aes(as.factor(T),EDUC)),
  ncol=2
  )

grid.arrange(
  ggplot(dta[dta$T==0,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(BLACK))) + 
    coord_polar(theta = "y") + 
    guides(fill=guide_legend(title="BLACK")),
  ggplot(dta[dta$T==1,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(BLACK))) +
    coord_polar(theta = "y") +
    guides(fill=guide_legend(title="BLACK")),
  ncol=2
)
grid.arrange(
  ggplot(dta[dta$T==0,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(HISP))) + 
    coord_polar(theta = "y") + 
    guides(fill=guide_legend(title="HISP"))+ylab("untreated"),
  ggplot(dta[dta$T==1,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(HISP))) +
    coord_polar(theta = "y") +
    guides(fill=guide_legend(title="HISP"))+ylab("treated"),
  ncol=2
)
```



```{r, echo=FALSE}
g <- ggplot(dta) + ylim(0,16) + xlab("Treatment (1,0)")
grid.arrange(
  g + geom_boxplot(aes(as.factor(T),RE74)),
  g + geom_boxplot(aes(as.factor(T),RE75)),
  g + geom_boxplot(aes(as.factor(T),RE78)),
  ncol=3
  )
g0 <- ggplot(dta[dta$T==0,]) +ylim(0,16) + xlab("Treatment (1,0)")
g1 <- ggplot(dta[dta$T==1,])+ ylim(0,7)  + xlab("Treatment (1,0)")
grid.arrange(
  g0 + geom_boxplot(aes(as.factor(T),RE74)),
  g0 + geom_boxplot(aes(as.factor(T),RE75)),
  g0 + geom_boxplot(aes(as.factor(T),RE78)),
  ncol=3)
grid.arrange(
  g1 + geom_boxplot(aes(as.factor(T),RE74)),
  g1 + geom_boxplot(aes(as.factor(T),RE75)),
  g1 + geom_boxplot(aes(as.factor(T),RE78)),
  ncol=3
  )
```



```{r}
dta %>% group_by(T) %>% summarise(mean=mean(RE78),sd=sd(RE78))
```
```{r}
t.test(dta$RE78~dta$T)
```


## Difference in means: pre treatment cov

```{r}
dta %>% group_by(T) %>% select(AGE,EDUC,MARR,BLACK,HISP,NODEGREE) %>% summarise_all(funs(mean(.)))
```
```{r}
lapply(c("AGE","EDUC","MARR","BLACK","HISP","NODEGREE"), function(x){t.test(dta[,x]~dta[,"T"])})
```

## Propensity score estimation


```{r}
summary(glm1 <- glm(T~.,   family = binomial(), data=dta))
```

```{r}
prs_df <- data.frame(pr_score = predict(glm1, type = "response"),
                     T = glm1$model$T)
sample_n(prs_df, size=5)
```

```{r, echo=FALSE}
grid.arrange(
  ggplot(prs_df[prs_df$T==0,])+geom_histogram(aes(x=pr_score))+ggtitle(""),
  ggplot(prs_df[prs_df$T==1,])+geom_histogram(aes(x=pr_score))
)
```

## Matching algorithm 

```{r}
formula <- formula(paste0("T~",paste0(colnames(dta)[!colnames(dta) %in% c("T","RE78")], collapse="+")))
mod_match <- matchit(formula, method="nearest",distance="logit", data=dta)
```

```{r}
dta.m <- match.data(mod_match)
dim(dta.m)
colnames(dta.m)
```


```{r, echo=FALSE}
fn_bal <- function(dta, variable) {
  dta$variable <- dta[, variable]
  support <- c(min(dta$variable), max(dta$variable))
  ggplot(dta, aes(x = distance, y = variable, color = as.factor(T))) +
    geom_point(alpha = 0.2, size = 1.3) +
    geom_smooth(method = "loess", se = F) +
    xlab("Propensity score") +
    ylab(variable) +
    theme_bw() +
    ylim(support)
}
grid.arrange(
fn_bal(dta.m, "AGE"),
fn_bal(dta.m, "EDUC"),
fn_bal(dta.m, "MARR"),
fn_bal(dta.m, "BLACK"),
fn_bal(dta.m, "HISP"),
fn_bal(dta.m, "NODEGREE"),
fn_bal(dta.m, "RE74"),
nrow=4)

```

```{r}
dta.m %>% group_by(T) %>% select(AGE,EDUC,RE74,MARR,BLACK,HISP,NODEGREE) %>% summarise_all(funs(mean(.)))
```

```{r}
lapply(c("AGE","EDUC","MARR","BLACK","HISP","NODEGREE"), function(x){t.test(dta.m[,x]~dta.m[,"T"])})
```

```{r}
t.test(dta.m$RE78~dta.m$T)
```


```{r, echo=FALSE}
grid.arrange(
  ggplot(dta.m) + geom_boxplot(aes(as.factor(T),AGE)),
  ggplot(dta.m) + geom_boxplot(aes(as.factor(T),EDUC)),
  ncol=2
  )

grid.arrange(
  ggplot(dta.m[dta.m$T==0,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(BLACK))) + 
    coord_polar(theta = "y") + 
    guides(fill=guide_legend(title="BLACK")),
  ggplot(dta.m[dta.m$T==1,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(BLACK))) +
    coord_polar(theta = "y") +
    guides(fill=guide_legend(title="BLACK")),
  ncol=2
)
grid.arrange(
  ggplot(dta.m[dta.m$T==0,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(HISP))) + 
    coord_polar(theta = "y") + 
    guides(fill=guide_legend(title="HISP"))+ylab("untreated"),
  ggplot(dta.m[dta.m$T==1,]) +
    geom_bar(aes(x=factor(1), fill=as.factor(HISP))) +
    coord_polar(theta = "y") +
    guides(fill=guide_legend(title="HISP"))+ylab("treated"),
  ncol=2
)
g <- ggplot(dta.m) + ylim(0,16) + xlab("Treatment (1,0)")
grid.arrange(
  g + geom_boxplot(aes(as.factor(T),RE74)),
  g + geom_boxplot(aes(as.factor(T),RE75)),
  g + geom_boxplot(aes(as.factor(T),RE78)),
  ncol=3
  )
g0 <- ggplot(dta.m[dta.m$T==0,]) +ylim(0,16) + xlab("Treatment (1,0)")
g1 <- ggplot(dta.m[dta.m$T==1,])+ ylim(0,7)  + xlab("Treatment (1,0)")
grid.arrange(
  g0 + geom_boxplot(aes(as.factor(T),RE74)),
  g0 + geom_boxplot(aes(as.factor(T),RE75)),
  g0 + geom_boxplot(aes(as.factor(T),RE78)),
  ncol=3)
grid.arrange(
  g1 + geom_boxplot(aes(as.factor(T),RE74)),
  g1 + geom_boxplot(aes(as.factor(T),RE75)),
  g1 + geom_boxplot(aes(as.factor(T),RE78)),
  ncol=3
  )
```




```{r}
lm_treat1 <- lm(RE78 ~ T, data = dta.m)
summary(lm_treat1)
```

```{r}
lm_treat1 <- lm(RE78 ~ T+EDUC+RE74, data = dta.m)
summary(lm_treat1)
```


## Package ATE

```{r}
# Y<-dta[,"RE78"]
# treat<-dta[,"T"]
# X<-dta[,!colnames(dta)%in%c("T","RE78","RE74","RE75","RE742","RE752","EDUC2","BLACKU74")]
# 
# fit1<- ATE(Y,treat,X)
# fit1
# summary(fit1)
# plot(fit1)
```


```{r}
summary(glm1 <- glm(T~.,   family = binomial(), data=dta))
dtanew <- dta
dtanew$p <- predict(glm1, type = "response")
#split_table <- read.csv(choose.files(), header=FALSE)
#dput(split_table)
split_table <- structure(list(V1 = 1:8, V2 = c(0.000591, 0.098016, 0.19544, 
0.390289, 0.585138, 0.779986, 0.877411, 0.926123), V3 = c(0.098016, 
0.19544, 0.390289, 0.585138, 0.779986, 0.877411, 0.926123, 0.974835
), V4 = c(1041L, 63L, 65L, 36L, 32L, 17L, 7L, 86L)), .Names = c("group", 
"lower", "upper", "obs"), class = "data.frame", row.names = c(NA, -8L
))

dtanew$group <- findInterval(dtanew$p, split_table$lower) 
table(dtanew$group)
split_table$obs
table(dtanew[,c("T","group")])
n <- data.frame(n=table(dtanew[,c("T","group")])[2,], group=0:8)
n[5,"n"]<-15
n[6,"n"] <- 13
n[9,"n"]<-7

dtanew %>% group_by(group, T) %>% select(AGE,EDUC,RE74,MARR,BLACK,HISP,NODEGREE) %>% summarise_all(funs(mean(.)))

for(i in sort(unique(dtanew$group))[-c(1,7,8,9)]){
  lapply(c("AGE","EDUC","MARR","BLACK","HISP","NODEGREE"), function(x){
  print(paste(x,t.test(dtanew[dtanew$group==i,x]~dtanew[dtanew$group==i,"T"])$p.value))}
    )
}

dtanew.matched <- dtanew[dtanew$T==1,]
for(i in sort(unique(dtanew$group))[-c(1,7,8)]){
dtanew.matched <- rbind(dtanew.matched, sample_n(dtanew[dtanew$group==i&dtanew$T==0,],size=n[n$group==i,"n"], replace=FALSE))
}

dtanew.matched %>% group_by(group, T) %>% select(AGE,EDUC,RE74,MARR,BLACK,HISP,NODEGREE) %>% summarise_all(funs(mean(.)))

summary(lm(RE78~T,data=dtanew.matched))
```


 
 